---
title: "style_notes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Text

I tried to use the mathematical notation for variables when we talk about them as abstract entities, e.g. "Here we will simulate a case where $N$ individuals each possess one of two mutually exclusive cultural traits. These alternative traits are denoted $A$ and $B$" or "We will call this proportion $p$. We could also track the proportion who possess trait $B$, but this will always be $1 - p$ given that the two traits are mutually exclusive".

However, when they are referred as part of the code, I used the code notation, e.g. "It then puts this proportion in the first slot of `p` in `output`". The distinction is not always very clear, e.g. in the first chapter the usage of $t_{\text{max}}$ versus `t_max`, but it seems reasonable to me.

Parts of the code, data structures, etc. are always in code notation (thus, no more **bold** for anything.)

For the time being, I kept the name of variables as they were. Some are descriptive (`population`, `trait`, etc.), some are not (`p`, `mu`, `N`). Perhaps a mix of the two is not that bad.

I used single quotes for technical terms: e.g. "The term 'oblique' means" or "In ggplot, 'geoms' describe what kind of visual representation should be plotted".

## Code

For the actual code, I mostly used the tidyverse [style guide](https://style.tidyverse.org) with some ad hoc choices. Name of functions and variables are with underscores to separate words and lowercase letters, e.g. `previous_population`, `biased_mutation`, etc.

If in the same chapter we have more than one function for the same model (for example because we add parameters), they are numbered as `unbiased_transmission_1()`, `unbiased_transmission_2()`, etc.

To avoid long lines (so that they print in a page), I separated function calls by argument per line (one or more):

```{r eval=FALSE}
output <- tibble(generation = rep(1:t_max, r_max),
                 p = as.numeric(rep(NA, t_max * r_max)),
                 run = as.factor(rep(1:r_max, each = t_max)))

population <- tibble(trait = sample(c("A", "B"), N, replace = TRUE,
                                    prob = c(p_0, 1 - p_0)))
```

Also to avoid long lines, I separated assignment after the assignment operator:

```{r eval=FALSE}
output[output$generation == 1 & output$run == r, ]$p <-
      sum(population$trait == "A") / N
```

For comments, I put them above the line commented (again, to avoid long lines). They start with capital letter, and do not have full stop.

```{r eval=FALSE}
# Determine 'mutant' individuals
mutate <- sample(c(TRUE, FALSE), N, prob = c(mu, 1 - mu), replace = TRUE)
```

For the rest, I followed the basic guidances:

* space after comma (not before).

* spaces surrounding operators (`-`, `<-`, `=`, `==`, etc.), but not `$`, `^`, and `:`.

* `<-` and not `=` for assignments.

* `TRUE` and `FALSE` instead of `T` and `F`.

* no space around parenthesis.

* `"` double quotes for quoting.

Probably I forgot something, we can add here.
