# Social network structure

For mathematical tractability models often assume well-mixed populations, where any individual is equally likely to interact with any other individual. These models often provide a good approximation to the real-world. Henrich's model, for example, has shown that a population's ability to maintain and accumulate cultural traits depends on its size, whereby larger populations are more likely to retain and improve more complex cultural traits than smaller ones. This model is a useful tool to study how population-level characteristics (here, demography) play a role in cultural dynamics. Given that individuals acquire information from other individuals through social learning, we might also suspect that individual-level phenomena are important, too. Commonly, individual interactions are represented as networks, whereby nodes (also referred to as vertex, _pl._ vertices) represent individuals, and ties (also referred to as edges) between any two nodes indicate a relationship (e.g. friendship or kinship, or willingness to cooperate, etc.). Networks can represent relationships between individuals, but just as well between countries, institutions, or any other kind of actors. 

In this chapter, we will cover the basics of social networks, how to create, analyse, and use them. Finally, we will use all of that to find an answer to the question: how do individual-level interaction patterns affect population-level cultural dynamics?

## Network basics
Let us start with a small group of colleagues who see each other on a regular base in the office, some are friends, and some are not (sometimes the friendships might not even be reciprocal). For this example, we use gossip as socially transmitted information. Say, we want to better understand how gossip is spreading through this group. At the simplest level, we might assume that: any two colleagues will exchange gossip whenever they meet. This would reduce the group of colleagues to a well-mixed population, where information can flow freely. However, assume that gossip is only transmitted if two friends interact. Now we have a situation, where information flow is constrained by the underlying friendship network. To explore, how information can spread in that friendship network, we first need to find a way to represent it in R. One way to represent networks is an adjacency matrix, $A$. It is a square matrix where all possible dyadic interactions of individuals in a population are represented as entries in its rows ($i$) and columns ($j$). Commonly, rows indicate interacting individuals, and columns are individuals who are recipients of this interaction. For example:

```{r 11.1}
m <- matrix(c(0,1,1, 0,0,1, 1,1,0), nrow = 3, byrow = TRUE)
row.names(m) <- c("A", "B", "C")
colnames(m) <- c("A", "B", "C")
m
```

Shown are the relationships between three individuals A, B, and C.  When we look at the first row, we see that A has no interaction with itself (indicated by the zero), but interacts both with B and C (indicated by the ones). From the next row, we see that B is only interacting with C, and the last row tells us that C interacting both with A and with B. This is an example for an asymmetric adjacency matrix because not all interactions are reciprocal (A is interacting with B but B is not interacting with A). We can test this by comparing the upper triangle of the matrix with the lower triangle of the matrix:
<!-- explain upper.tri and lower.tri -->
```{r 11.2}
all(m[upper.tri(m)] == m[lower.tri(m)])
```

We can make this matrix symmetric by replacing the lower matrix triangle with the upper one (this requires us to first transpose the triangle using the `t()` function):

```{r 11.3}
# Either
m[lower.tri(m)] <- t(m)[lower.tri(m)]
# or use m[2,1] <- 1
m
# Check whether the matrix is symmetric now:
all(m[upper.tri(m)] == m[lower.tri(m)])
```

Now let us turn back to our group of co-workers, and let us create a random friendship network. We start with an empty adjacency matrix, add friendships to the upper triangle and then copy the mirror image to the lower triangle (this is assuming that friendships are reciprocal):

```{r 11.4}
n <- 10
adjm <- matrix(NA, ncol = n, nrow = n)
adjm
```

Now, we randomly assign friendships to one of the triangles. We sample from zeros (no friends) and ones (friends) with a probability that indicates how many friendships there are in the network (`friendschipRatio`):

```{r 11.5}
friendshipRatio <- 0.5

adjm[upper.tri(adjm)] <- sample(x = 0:1, 
                                size = n*(n-1)/2, 
                                replace = TRUE, 
                                prob = c(1-friendshipRatio, friendshipRatio))
adjm

# add the mirror image to the lower triangle
adjm[lower.tri(adjm)] <- t(adjm)[lower.tri(adjm)]
adjm

# replace diagonal (these indicate self-references) with zeros
diag(adjm) <- 0
adjm
```


This is all we need to describe the interactions and the flow of information within this group. And while we could already use other tools in R to analyse it, we should first try to visualise this network to get a better intuition of what this group looks like. The `igraph` package provides many incredibly useful functions and tools to create, work with, and analyse networks. We will rely on many of its functions in this chapter. 

## Plotting networks 
Given that we already have an adjacency matrix, we can create a network using the `graph_from_adjacency_matrix()` function. 

```{r 11.6, message=FALSE}
library(igraph)
net <- graph_from_adjacency_matrix(adjm)
net
```

When we return the `net` object, we receive a lot of information about our network, for example, that it is there are `r vcount(net)` vertices and `r ecount(net)` edges (for more information be sure to have a look at this short [igraph introduction](https://igraph.org/r/doc/aaa-igraph-package.html)). However, we did not receive an actual network figure. For this, we can use `R`'s generic plot function:

```{r 11.7, fig.cap="A simple network with 10 vertices, which are connected by edges (arrows)."}
set.seed(1)
plot(net)
```

This is the most basic network plot where each node (with the numbers 1 to `r n`) and their connections (edges/ties) are plotted such that nodes that receive more connections are more central and those that receive less are more peripheral. Note, every time you plot the network, it will have a slightly different layout. If you would like to preserve the node positions you can uncomment the `set.seed()` function in the code chunk above.

There are many ways to change the looks of network plots. The igraph manual is an ideal starting point to learn more about it. For example, we could plot nodes based on specific layout functions:

```{r 11.8, fig.cap="Example for two different network layouts, the grid and the ring."}
par(mfrow = c(1,2))
plot(net, layout = layout.grid(net), main = "Grid layout")
plot(net, layout = layout.circle(net), main = "Ring layout")
```

Or we can directly change the size, and colour of nodes:

```{r 11.9, fig.cap="Using syling arguments with the `plot()` function allows to manipulate the appearence of vertices, edges, and labels."}
par(mfrow = c(1,1))
set.seed(1)
plot(net, 
     vertex.color = "dodgerblue", 
     vertex.label.color = "white",
     vertex.size = 20,
     edge.color = "black",
     edge.width = 1,
     edge.arrow.size = 0.5,
     main = "Default layout with styling")
```

In the previous example, we have changed the colour of the nodes' background and text, and their overall size (attributes starting with `vertex.`). We have also changed the colour of the edges, their width, and the size of the tips (attributes starting with `edge.`).

Let's say, we know the age of each individual and we would like to visualise age as different colours of the nodes. For that, we should first store the colour information for each vertex (using the `V()` function) and then change the vertex colour in the plot function:

<!-- explain hear.colors -->

```{r 11.10}
# There is no age attribute set
V(net)$age

# Add an age attribute and randomly assign an age class
V(net)$age <- sample(x = 1:10, size = n, replace = TRUE)

# The network now has an age attribute
V(net)$age

V(net)$colour <- heat.colors(n = 10, rev = TRUE)[V(net)$age]

net
```

The `net` object now has two attributes (one is called age, with numeric values, and one is called colour, with character values). We can now plot this graph:

```{r 11.11, fig.cap="In this plot we indicate a node's age by colouring the node with a different colour (pale yellow: younger, red: older)"}
set.seed(1)
plot(net, 
     vertex.color = V(net)$colour, 
     vertex.label.color = "black",
     vertex.size = 20,
     edge.color = "black",
     edge.width = 1,
     edge.arrow.size = 0.5)
```

It can be useful to remove the labels from the nodes and reduce the node size, especially when the networks become large. Also, given that we have a symmetric network (all relationships are reciprocal), we can get rid of the arrow tips. We do this by telling igraph that the network is undirected.
<!-- explain as.undirected -->
```{r 11.12, fig.cap="We can remove vertex labels to reduce clutter."}
net <- as.undirected(net)
set.seed(1)
plot(net, 
     vertex.color = V(net)$colour, 
     vertex.label = "",
     vertex.size = 9,
     edge.width = 1,
     edge.arrow.size = 0.5)
```

One final styling example. So far, we have the edges indicate the existence of friendship relationships, which is a binary quality. However, we can also imagine friendship as a continuous variable (from strong = 1 to none = 0). Using the edge function `E()`, we can add attributes to the edges in the network. We can use those to alter the width of edges depending on their relationship strength. To do that, we will first change the adjacency matrix (from binary to continuous values) and then plot the new network.
<!-- explain runif() -->
```{r 11.13}
# create a copy of the adjacency matrix
adjm2 <- adjm

# draw random values from a uniform distribution and replace the matrix entries that have a value of 1
adjm2[upper.tri(adjm2)] <- runif(n = sum(upper.tri(adjm2))) * adjm2[upper.tri(adjm2)]

# copy the mirror image to the lower triangle
adjm2[lower.tri(adjm2)] <- t(adjm2)[lower.tri(adjm2)]

# create the network
net2 <- graph_from_adjacency_matrix(adjm2, weighted = TRUE, mode = "undirected")
net2
```

Note, igraph automatically added an attribute called `weight`. Now, let's plot this network with the edge width representing their weight:

```{r 11.14, fig.cap="To indicate different strength in the relationship of two nodes, we can vary the width of edge between them."}
V(net2)$age <- V(net)$age
V(net2)$colour <- V(net)$colour

plot(net2, 
     vertex.color = V(net2)$colour,
     vertex.label = "",
     vertex.size = 9,
     edge.width = E(net2)$weight*5,
     edge.arrow.size = 0.5)
```

You can now observe strong and weak relationships between individuals, their location relative to each other, and how they cluster. In the next section, we will try to quantify these observations.

## Analyse social networks
There is a variety of measures to describe the structure of our graph (another term for network). We can generally differentiate between properties of the network and properties of the vertices:

### Network properties 
We can retrieve the most basic information about our network using:
```{r 11.15}
# vertex information
V(net2)

# edge information
E(net2)
```

If our networks have attributes stored (for the vertices or edges), we can retrieve them with the following two functions: 
```{r 11.16}
# vertex attributes
get.vertex.attribute(graph = net2)

# edge attributes
get.edge.attribute(net2)
```

With `vcount()` and `ecount()`, we can return the number of vertices (`r vcount(net)`) and edges (`r ecount(net)`) in our graph. 
```{r 11.17}
# vertex count
vcount(net)

# edge count
ecount(net)
```

Let us now look at a variety of measures that we can calculate to characterise our network.

**Diameter** is a measure for the longest (geodesic) path, i.e. the largest number of steps that are necessary to reach two vertices in a network (using `farthest_vertices()` we can return the ID of the two vertices).
```{r 11.18}
diameter(graph = net)
```

**Average path length** is the average number of steps that need to be traversed between any two vertices (aka as dyad). We can also use the `distance()` function to return a distance matrix similar to the adjacency matrix.
```{r 11.19}
# average distance between any dyad
mean_distance(graph = net)
```

**Edge density** is the proportion of edges present in the graph relative to the number of possible edges (i.e. in a fully connected network with the same number of nodes).
```{r 11.20}
edge_density(graph = net)
```

**Reciprocity** (relevant for directed graphs only) calculates the proportion of mutual edges. As we have a directed graph, this value is one. 
```{r 11.21}
reciprocity(graph = net)
```

**Clustering coefficient** (aka transitivity, or cliquishness) is the probability that the two neighbours of a vertex are neighbours of each other. This is also called a triangle, and often the image of 'my friends are friends with each other' is used.
```{r 11.22}
transitivity(graph = net)
```


### Vertex properties
Additional to these high-level measures, we can use a series of node-level measures to describe connectivity in more detail:

**Degree centrality** refers to the number of (incoming/outgoing/both) edges of a vertex. We can use the `degree()` function to determine the degree centrality of each node:
<!-- explain identical() -->
```{r 11.23}
# number of edges that connected with each node
degree(graph = net)

# as we have an undirected network, the count of incoming and outgoing edges is identical
identical(degree(graph = net, mode = "in"), 
          degree(graph = net, mode = "out"))

# we can use the mean of all degree centralities as a general measure for the entire network
mean(degree(graph = net))
```

**Strength** is similar to degree centrality but relevant for weighted networks. It is the sum of all adjacent edge weights (a node might have many edges but with very low weights and so with high degree centrality but low strength). In our case (unweighted network), `degree()` and `strength()` produce the same result:
```{r 11.24}
sort(strength(graph = net))
```

**Closeness centrality** represents the number of steps it takes from a given vertex to any other vertex in the network. It is a measure for how long information on average takes to arrive at this node. 
```{r 11.25}
closeness(graph = net)
```
Note that the values are $<$1. This is because `igraph` defines closeness centrality as 'the inverse of the average length of the shortest paths to/from all the other vertices in the graph.'

**Betweenness centrality** is the number of shortest paths between nodes that pass through a particular node. It is often seen as a measure for a node's gatekeeping or brokerage potential:
```{r 11.26}
betweenness(graph = net)
```

**Eigenvector centrality** is the eigenvector of the adjacency matrix. Vertices with a high eigenvector centrality are connected to many individuals who are connected to many individuals, and so on (see also page rank, `page_rank()`, and authority, `authority_score()`, score functions).
```{r 11.27}
eigen_centrality(graph = net)$vector
```

## Using social networks to model information transmission
Now that we know how to generate, plot, and analyse networks: how can we use them in the context of information sharing. Or to come back to our initial example, the spread of gossip? Let us start with a very simple model: we initiate the simulation be endowing one individual with a piece of information (or gossip). We then simulate how many time steps it will take until the information has spread through the entire network. 

### Gossip transmission on networked populations
```{r 11.28}
# set up vector that stores whether an individual possesses the latest gossip:
gossip_meter <- rep(FALSE, n)
gossip_meter

# choose a random individual to start gossiping
gossip_meter[sample(x = n, size = 1)] <- TRUE
gossip_meter

# choose an individual with gossip
if(sum(gossip_meter) == 1){
  gossip_provider <- which(gossip_meter)
} else {
  gossip_provider <- sample(x = which(gossip_meter), size = 1)
}

# choose a neighbour in the network
neighbours <- which(adjm[gossip_provider,] == 1)

# only execute if there is a neighbour
if(length(neighbours) > 0){
  # choose one neighbour if several are present
  if(length(neighbours) == 1){
    neighbour <- neighbours
  } else {
    neighbour <- sample(x = neighbours, size = 1)
  }
  # update gossip indicator of the neighbour
  gossip_meter[neighbour] <- TRUE
}
gossip_meter
```

This is already sufficient to simulate the spread of gossip on a network. In the next step, let us loop over these instructions and record the number of time steps it takes until everyone received the gossip. But first, we should write a different `sample()` function, as the generic function misbehaves with samples of length equal to one. We will use the following: 

```{r 11.29}
mySample <- function(x, prob = NULL){
  if(length(x) == 1){
    return(x)
  } else {
    return(sample(x = x, size = 1, prob = prob))
  }
}
```

Now we can use this simpler version of the sampling function (which is directly tailored to our needs), replace it in the code above, and add a `for` loop: 

```{r 11.30}
# total number of turns to loop over
Turns <- 50

# a matrix to record how many individuals have the gossip at time x
res <- tibble(turn = 1:Turns, gossipers = rep(0, Turns))

# set up gossip indicator
gossip_meter <- rep(FALSE, n)

# choose a random individual to start gossipping
gossip_meter[sample(x = n, size = 1)] <- TRUE

for(turns in 1:Turns){
  res[turns,2] <- sum(gossip_meter)/n
  # choose an individual with gossip
  gossip_provider <- mySample(which(gossip_meter))
  
  # choose a neighbour in the network
  neighbours <- which(adjm[gossip_provider,] == 1)
  
  # only execute if there is a neighbour
  if(length(neighbours) > 0){
    # choose one neighbour ...
    neighbour <- mySample(neighbours)
    # and update its gossip indicator
    gossip_meter[neighbour] <- TRUE
  }
}

# have a look at the time course
res$turn
```


```{r 11.301, fig.cap="Gossip spreads first quickly and then more slowly throughout a population."}
# plot result
library(ggplot2)
ggplot(res) + 
  geom_line(aes(x = turn, y = gossipers)) +
  ylab("prop. of pop. with gossip") +
  theme_bw()
```

Let us adapt the code so that we can run it repeatedly, to get a better feeling for the average amount of time it takes for the information to spread. 

```{r 11.31}
gossip_model_1 <- function(ADJM, TURNS, SIM = 1){
  n <- nrow(ADJM)
  # a matrix to record how many individuals have the gossip at time x
  res <- tibble(turn = 1:TURNS, gossipers = rep(0, TURNS))
  # set up gossip indicator
  gossip_meter <- rep(FALSE, n)
  # choose a random individual to start gossiping
  gossip_meter[sample(x = n, size = 1)] <- TRUE
  
  for(turns in 1:TURNS){
    res[turns,2] <- sum(gossip_meter)/n
    # choose an individual with gossip
    gossip_provider <- mySample(which(gossip_meter))
    # choose a neighbour in the network
    neighbours <- which(ADJM[gossip_provider,] == 1)
    # only execute if there is a neighbour
    if(length(neighbours) > 0){
      # choose one neighbour ...
      neighbour <- mySample(neighbours)
      # and update its gossip indicator
      gossip_meter[neighbour] <- TRUE
    }
  }
  res$simulation <- SIM
  return(res)
}
```

With this function of our model, we can easily re-run the simulation several times and receive the time trajectories for each simulation:

```{r 11.32, cache=TRUE}
data <- do.call("rbind", lapply(1:100, function(sim){
  gossip_model_1(ADJM = adjm, TURNS = 100, SIM = sim)
}))
```

Here, we are using an additional new function, `do.call()`. This is a wrapper function that repeatedly applies a function to the individual elements of an object and then returns the result. In this case, we want to bind all the rows (therefore using `ribnd()`) of the list object that `lapply()` returns. `do.call()` takes the first list element and applies `rbind()` to it and the second element. In the next step it takes the resulting (now longer) data set and binds the rows of the next list element, and so on. After binding all the rows together, we can use the returned `data` object and plot it using `ggplot`. 

```{r 11.33, fig.cap="Repeated runs of the gossip simulation shows the asymptotic spread of gossip in a population."}
ggplot(data) + 
  geom_line(aes(x = turn, y = gossipers, group = simulation), alpha = .2) +
  ylab("prop. of pop. with gossip") +
  theme_bw()
```

We can see that the initial spread is very fast and then tapers off. At the beginning there are a lot of individuals who do not have the gossip and transmission will be successful. However, the more individuals with the gossip there are, the less likely it is to find someone who does not yet have the gossip. How does this result compare to simulations with a fully connected network (i.e. where every individual is equally likely to interact with any other individual)?

To make the creation of networks easier, let us write down a function to generate networks as we have done before:

```{r 11.34}
create_network <- function(N, FRIENDSHIPRATIO){
  adjm <- matrix(0, ncol = N, nrow = N)
  adjm[upper.tri(adjm)] <- sample(x = 0:1, 
                                  size = N*(N-1)/2, 
                                  replace = TRUE, 
                                  prob = c(1-FRIENDSHIPRATIO, FRIENDSHIPRATIO))
  adjm[lower.tri(adjm)] <- t(adjm)[lower.tri(adjm)]
  return(adjm)
}
```

Now we can create a fully connected network, run the gossip model on it, and plot the results:
<!--recall how lapply function works form chapter 10-->
```{r 11.35, cache=TRUE}
adjm_full <- create_network(N = n, FRIENDSHIPRATIO = 1)

data_full <- do.call("rbind", 
                     lapply(1:100, function(sim){
                       gossip_model_1(ADJM = adjm_full, TURNS = 100, SIM = sim)
                       })
                     )
```

<!--explain geom_smooth-->

```{r 11.36, fig.cap="Gossip spreads faster in a fully connected network, compared to a less connected network.", message=FALSE}
data$type <- "colleague network"
data_full$type <- "fully connected"
data_combined <- rbind(data, data_full)

ggplot(data_combined) + 
  geom_point(aes(x = turn, y = gossipers, color = type), alpha = .05) +
  geom_smooth(aes(x = turn, y = gossipers, color = type)) +
  ylab("prop. of pop. with gossip") +
  theme_bw()
```

This plot shows that the fully connected network reaches saturation a little bit faster than the colleague network. The difference becomes even more obvious when we re-run this simulation with larger networks:

```{r 11.37, cache=TRUE}
adjm_colleagues <- create_network(N = 25, FRIENDSHIPRATIO = .20)
adjm_full <- create_network(N = 25, FRIENDSHIPRATIO = 1)

data_colleagues <- do.call("rbind", lapply(1:100, function(sim){
  gossip_model_1(ADJM = adjm_colleagues, TURNS = 100, SIM = sim)
}))

data_full <- do.call("rbind", lapply(1:100, function(sim){
  gossip_model_1(ADJM = adjm_full, TURNS = 100, SIM = sim)
}))
```

```{r 11.38, fig.cap="When networks are larger, gossip spreads even faster in fully connected networks. Note that there are some simulations where gossip does not spread (red dotes at the bottom). This happens when a simulation begins with an individual, who is in possession of gossip, but does not share a connection with any other individual in the population.", message=FALSE}
data_colleagues$type <- "colleague network"
data_full$type <- "fully connected"
data_combined2 <- rbind(data_full, data_colleagues)

ggplot(data_combined2) + 
  geom_point(aes(x = turn, y = gossipers, color = type), alpha = .05) +
  geom_smooth(aes(x = turn, y = gossipers, color = type)) +
  ylab("prop. of pop. with gossip") +
  theme_bw()
```



### How does network structure affect information transmission?
While the previous example has shown us the difference in the transmission speed of information in a structured versus an unstructured population (i.e. the fully connected network which is equivalent to a well-mixed population), we still do not know how network characteristics affect the transmission of information. To get a better understanding of it, we will simulate the transmission of information in networks of different sizes and connectivity. Here, we will simulate Small-World networks (also known as Watts-Strogatz graphs). These are networks that are characterised by high clustering and short average path length. 

For this type of analysis we do not need to record each time step, instead, we can let our function return the time step at which, e.g. 75\% of the population have heard the gossip:

```{r 11.39}
gossip_model_2 <- function(ADJM, PROP, TURNS, SIM = 1){
  n <- nrow(ADJM)
  # set up gossip indicator
  gossip_meter <- rep(F, n)
  # choose a random individual to start gossipping
  gossip_meter[sample(x = n, size = 1)] <- TRUE
  
  timestep <- NA
  for(turns in 1:TURNS){
    # record the time step when a defined proportion of the population has the gossip
    if(sum(gossip_meter)/n >= PROP & is.na(timestep)){
      timestep <- turns 
    }
    # choose an individual with gossip
    gossip_provider <- mySample(which(gossip_meter))
    # choose a neighbour in the network
    neighbours <- which(ADJM[gossip_provider,] == 1)
    # only execute if there is a neighbour
    if(length(neighbours)>0){
      # choose one neighbour ...
      neighbour <- mySample(neighbours)
      # and update its gossip indicator
      gossip_meter[neighbour] <- TRUE
    }
  }
  # if PROP is not reached, return the maximum value
  if(is.na(timestep)){
    timestep <- TURNS
  }
  return(timestep)
}
```

By default, this function will return the maximum turn number, if the set target proportion is not reached (however, we could also let the function return a different value or a character expression, such as `MAX`).

```{r 11.40}
adjm <- create_network(N = 50, FRIENDSHIPRATIO = .25)
gossip_model_2(ADJM = adjm, PROP = .75, TURNS = 100, SIM = 1)
```
When you run d

Now, let us run these functions for different group sizes and different connectivities, and let us repeat these simulations several times to get a better estimate:
<!--recall how expand.grid works-->
```{r 11.41, cache=TRUE}
parameters <- expand.grid(groupSize = c(10,50),#c(10,30,50,70,90),
                          reconnections = c(0,.01,.05,.1,.5,1),
                          repetition = 1:50)
tmp <- lapply(1:nrow(parameters), function(p){
  net <- watts.strogatz.game(dim = 1, size = parameters[p,"groupSize"], nei = 2, 
                             p = parameters[p,"reconnections"], loops = FALSE, 
                             multiple = FALSE)
  adjm <- get.adjacency(net)
  model <- gossip_model_2(ADJM = adjm, PROP = .5, TURNS = 500, 
                          SIM = parameters[p,"repetition"])
  tibble(time = model, 
             degree = mean(degree(net)), 
             path = average.path.length(net))
})

tmp2 <- do.call("rbind", tmp)
parameters <- cbind(parameters, tmp2)
```

To plot the average time of our repeated simulations, we need to first calculate this average:
<!--explain what is happening here-->
```{r 11.42}
library(tidyverse)
parameters %>% 
  group_by(groupSize, reconnections) %>% 
  summarise(meanTime = mean(time), 
            meanDegree = mean(degree),
            meanPath = mean(path)) -> data
```

Let us have a look at both the average network degree and the average path length of the networks:
<!--explain why you use factor()-->
```{r 11.43, fig.cap="Based on how small world networks are created, the average network degree remains unchanged even if the probability of reconnection increases. Note that we use the plotting function `geom_jitter()` which works just like `geom_point()` but is also adding a little y-axis jitter to the points so that we can see overlapping points."}
parameters$groupSize <- factor(parameters$groupSize)

ggplot(parameters, aes(x = reconnections, y = degree, col = groupSize)) + 
  geom_jitter(height = .01, alpha = .5) +
  scale_y_continuous(limits = c(3.5,4.5)) +
  xlab("prob. of reconnection") +
  ylab("average degree") +
  theme_bw()
```

Note, the average network degree is constant at $k =$ `r mean(parameters$degree)`. This is due to the way Small World Networks are created. We start with a regular lattice network and then randomly select an edge and change the vertex on one end (e.g. from A--B to A--C). Thus, while some individuals might now have fewer (or even zero) edges and others might have more, the average number of edges and so the average degree remains fixed an independent of the number of rewired edges.

```{r 11.44, fig.cap="In larger populations average path length is longer when reconnection prbability is low. For smaller populations we do not see this effect at all."}
ggplot(parameters, aes(x = reconnections, y = path, col = groupSize)) + 
  geom_point() + 
  xlab("prob. of reconnection") +
  ylab("average path length") +
  theme_bw()
```

This, however, is not true for the path length. Instead, we find that larger networks (with otherwise the same network parameters) have longer average path lengths (i.e. it takes on average more steps to get from one to another individual). Also, as we increase the probability to reconnect edges average path length goes down. This is because random connections might connect distant sections of the network and so drastically lower the number of steps required to get from one individual to another. 

```{r 11.45, fig.cap="In larger populations, and especially in those with longer average path length, it takes longer for information to spread than in smaller populations or those with shorter average path length."}
ggplot(parameters, aes(x = path, y = time, col = groupSize)) + 
  geom_point(alpha = .33) + 
  xlab("average path length") +
  ylab("time") +
  theme_bw()
```

From this graph, we can see that the spread of information decreases as networks become larger and as average path length increases. 


### Complex versus simple contagion

The spread of information is not only affected by the network shape but also by the manner of information transmission. That is, often information is not simply transmitted from one individual to another in a simple contagion like manner but instead requires increased social facilitation. This can both be repeated exposure to the information or a certain proportion of social partners who possess this information. In other words, often we are more likely to acquire behaviours from others if this behaviour is more frequent in our neighbourhood. This kind of transmission is called complex contagion (see e.g. @centola_spread_2010). But how does simple and complex contagion affect the spread of information? To model this, we can write a small model where we randomly select individuals and check whether they acquire information from their social contacts based on a threshold value, i.e. the number of contacts required to acquire information. 

Let us start with a model of simple contagoin (requires only one social contact that possesses a new beahviour or information to acquire it):

```{r 11.46, cache=TRUE}
informationSpread <- function(ADJM, MODE = "simple", ROUNDS = 100, CONTACTS = 1){
  N <- nrow(ADJM)
  
  # set up behaviour
  behaviour <- rep(F, N)
  behaviour[sample(N,size = 10)] <- TRUE
  
  # recording data
  fraction <- rep(0, ROUNDS)
  
  # loop
  for(round in 1:ROUNDS){
    # record
    fraction[round] <- sum(behaviour)/N
    
    # pick random individual
    focal <- sample(x = N, size = 1)
    # who are its neighbours
    neigh <- adjm[focal, ] == 1
    
    if(sum(behaviour[neigh]) >= CONTACTS){
      behaviour[focal] <- TRUE
    }
    
  }
  return(tibble(mode = MODE, 
                frac = fraction, 
                round = 1:ROUNDS))
}

# Simulations with clustered (lattice) networks
net <- watts.strogatz.game(size = 10, dim = 2, nei = 2, p = .01)
adjm <- as.matrix(get.adjacency(net))
clu <- informationSpread(ADJM = adjm, MODE = "simple", ROUNDS = 1000, CONTACTS = 1)
clu$network <- "clustered"

# Simulations with random (small world) networks
net2 <- erdos.renyi.game(n = 100, p.or.m = 600, type = "gnm", directed = FALSE, loops = FALSE)
adjm2 <- as.matrix(get.adjacency(net2))
rand <- informationSpread(ADJM = adjm2, MODE = "simple", ROUNDS = 1000, CONTACTS = 1)
rand$network <- "random"

res <- rbind(clu, rand)
```


```{r 11.461, fig.cap="When information spreads in a simple contagion manner, we find only small differences spread in clustered and random networks. In both cases information spreads quickly at first and then slows down as the population reaches saturation."}
ggplot(res) + 
  geom_line(aes(x = round, y = frac, col = network, linetype = mode), size = .9) + 
  theme_bw()
```

For the simple contagion case we find very little difference between the two network types. Let us now add the complex contagion case to the simulation:
<!--explain how do.call() works-->
```{r 11.47, cache=TRUE}
# for repeated interactions 
informationSpread <- function(NET, MODE = "simple", ROUNDS = 100, CONTACTS = 1){
  if(NET == "clustered"){
    net <- watts.strogatz.game(size = 10, dim = 2, nei = 2, p = .01, loops = FALSE, multiple = FALSE)
  } else if(NET == "random"){
    net <- erdos.renyi.game(n = 100, p.or.m = 600, type = "gnm", directed = FALSE, loops = FALSE)
  }
 adjm <- as.matrix(get.adjacency(net)) 
 
  N <- nrow(adjm)
  
  # set up behaviour
  behaviour <- rep(F, N)
  behaviour[sample(N,size = 10)] <- TRUE
  
  # recording data
  fraction <- rep(0, ROUNDS)
  
  # loop
  for(round in 1:ROUNDS){
    # record
    fraction[round] <- sum(behaviour)/N
    
    # pick random individual
    focal <- sample(x = N, size = 1)
    # who are its neighbours
    neigh <- adjm[focal, ] == 1
    
    if(sum(behaviour[neigh]) >= CONTACTS){
      behaviour[focal] <- TRUE
    }
    
  }
  return(tibble(network = NET, 
                mode = MODE, 
                frac = fraction, 
                round = 1:ROUNDS))
}

# repeated runs
reps <- 10

# simulte complex and simple contagion for clustered networks
set.seed(11)
clu_simple <- do.call("rbind", 
                      lapply(1:reps, function(x) informationSpread(NET = "clustered", 
                                     ROUNDS = 1000, CONTACTS = 1, MODE = "simple")))
clu_comp <- do.call("rbind", 
                    lapply(1:reps, function(x) informationSpread(NET = "clustered", 
                                     ROUNDS = 1000, CONTACTS = 4, MODE = "complex")))
clu <- rbind(clu_simple, clu_comp)

# simulte complex and simple contagion for random networks
rand_simple <- do.call("rbind", 
                       lapply(1:reps, function(x) informationSpread(NET = "random", 
                                      ROUNDS = 1000, CONTACTS = 1, MODE = "simple")))
rand_comp <- do.call("rbind", 
                     lapply(1:reps, function(x) informationSpread(NET = "random", 
                                      ROUNDS = 1000, CONTACTS = 4, MODE = "complex")))
rand <- rbind(rand_simple, rand_comp)

# combine results
res <- rbind(clu, rand)
```

```{r 11.48, fig.cap="While we find almost no difference in the spread of information between clustered and random networks when information is spread in a simple contagion manner, there are larger differences when the spread follows complex contagion dynamics. Here, we find faster spread of information in clustered networks. ", message=FALSE}
ggplot(res) + 
  stat_smooth(aes(x = round, y = frac, col = network, 
                  linetype = mode), size = .9, se = T) +
  ylab("Proportion of individuals with information") +
  xlab("Round") + 
  theme_bw()
```

This figure tells us that, while there is no apparent difference in the spread of information in clustered and random networks for simple contagion, we find that information spreads faster in clustered networks if the transmission is akin to complex contagion. The reason for this is that in clustered networks an individual's neighbours are more likely to also be connected. This increases the likelihood that the neighbours also share the same information, and, in turn, increases the individual's exposure to this information. 



## Summary of the model
In this chapter we have explored individual-level effects on population-level outcomes. That is, how the structure of individual interactions affect the spread of information in a population. We have seen that both network characteristics (degree and path length) but also the mode of information transmission (simple versus complex) can have strong effects on how efficiently information travels through a population.

## Further Reading
There is an increasing number of theoretical models that is looking at the effect of network characteristics on the spread of cultural traits, e.g. @osullivan_mathematical_2015. Several empirical studies with humans (@hill_hunter-gatherer_2014) and non-human animals (@aplin_social_2012), have recorded network structures and reported on the effects of network structure on the spread of novel information. It is also interesting to ask, how the network structure itself might be the result of cultural dynamics (see e.g. @smolla_cultural_2019). A good overview over relevant literature is provided in a review by @derex_cumulative_2020. 



